Temporality is Important for Networks of Hospital Patient Transfers

### 2024-11-04 reports/report.qmd
yep report updates

---
title: "Temporality is Important for Networks of Hospital Patient Transfers"
authors:
    - name: David Wu
    - name: Michael Lydeamore
    - name: Andrew Stewardson
    - name: Tjibbe Donker
    - name: Nic Geard
    - name: "******"
date: last-modified
format:
    html: default
    pdf:
        include-in-header:
            - text: |
                \usepackage{mathtools}
        papersize: a4
bibliography: refs.bib
csl: plos.csl
jupyter: python3
filters:
    - diagram
execute:
    echo: false
    warning: false
    error: false
---

@@

@@

## Explicitly Temporal Models
Previous analyses of movement and spread of contagions through these networks have made the assumption that the weight $w_{ij}$ for a given edge does not change over time, i.e. the network is static.
One advantage that the VAED provide us is the minute-scale granularity of patient admissions, which allows us to construct temporal networks and models of patient movement.

@@

Patient transfers can be considered as edges from some source hospital $x_i$ at some time $t_m$ to a target hospital $x_j$ at some time $t_n$. We could work with the exact times, but this will lead to an overly sparse representation that is difficult to work with.
Instead, we discretise time into bins by treating movements involving some arbitrary hospital $x_i$ between times $t_n$ and $t_n + W$ as instead involving a single node at $v_{i,n} \coloneqq (x_i, t_n)$. 
Then we can have a temporal network that can be treated as a static network as in @sec-network-modelling, i.e. we have edge weights $w_{ij}$ along the edge $e_{ij} \coloneqq (v_i, v_j)$.

@@

We define the _extent_ of a simulation by the number of facilities that some patient has moved to or through within a given number of days. We see in @fig-extent-30 that we get starkly different distriubtions of extent for the naïve static model, commpared to the other models.

@@

We can frame the empirical cumulative distribution (ECDF) of hitting times from a given seed hospital to a target hospital in a survival analysis view, and compare different experimental setups by comparing the ECDFs of the combinations of hitting times with log-rank tests. Each pair of hospitals yields a separate log-rank test statistic.
Using a crude Bonferroni correction to counteract the multiple comparisons (which occur due to the $N=338^2$ combinations of seed and target hospitals), we find that there are significant differences between each type of model. We note spikes at very low p-values when comparing the naïve static model to other models.

@@

fig, axs = plt.subplots(nrows=NN, ncols=NN, squeeze=True, sharex=True, sharey=True)

@@

    axs[index].axvline(confidence_threshold, color='r')
    axs[index].set_yscale('log')
    sns.despine(ax=axs[index])

@@

>>>>

---
title: "Temporality is Important for Networks of Hospital Patient Transfers"
authors:
    - name: David Wu
    - name: Michael Lydeamore
    - name: Andrew Stewardson
    - name: Tjibbe Donker
    - name: Nic Geard
    - name: "******"
date: last-modified
format:
    html: default
    pdf:
        include-in-header:
            - text: |
                \usepackage{mathtools}
        papersize: a4
bibliography: refs.bib
csl: plos.csl
jupyter: python3
# filters:
#     - diagram
execute:
    echo: false
    warning: false
    error: false
---

@@

# Glossary

| symbol | name | description |
|--------|------|-------------|
| v | vertex | |
| e | edge | |
| i | index of vertex | |
| j | index of other vertex | |
| t | time | |
| x | location | hospital |
| W | bin width | discretisation of the temporal network |
| $\gamma$ | discharge | rate of discharge |
| N | num | size of something |
| k | hitting time | (expected) |
| $\lambda$ | rate | of an event / process |
| T | time period | of the _entire_ simulation |

define a vertex $v_i = (x_i, t_i)$ made up of a location and time resp.

@@

## Explicitly Temporal Models
Previous analyses of movement and spread of contagions through these networks have made the assumption that the weight $w_{ij}$ for a given edge does not change over time, i.e. the network is static.
This implies that the rate movement of patients between hospitals is Poisson, or that the time between successive admissions for a given patient is exponentially distributed. Empirically, this is poor modelling choice, since the readmission window, as shown in @fig-readmission-window, is definitely not exponentially distributed, and has a significant heavy tail. A naïve static approximation that matches the mean readmission window will be inaccurate; in the following, we will attempt to resolve this problem in three increasingly granular ways:

1. using a simple piecewise exponential approximation, by choosing a single threshold
2. using a series of static networks with ...
3. using a temporal network that exactly captures the readmission window distribution

<!-- One advantage that the VAED provide us is the minute-scale granularity of patient admissions, which allows us to construct temporal networks and models of patient movement. -->

@@

Patient transfers can be considered as edges from some source hospital $x_i$ at some time $t_i$ to a target hospital $x_j$ at some time $t_j$. We could work with the exact times, but this will lead to an overly sparse representation that is difficult to work with.
Instead, we discretise time into bins by treating movements involving some arbitrary hospital $x_i$ between times $t_i$ and $t_i + W$ as instead involving a single node at $v_i \coloneqq (x_i, t_i)$. 
Then we can have a temporal network that can be treated as a static network as in @sec-network-modelling, i.e. we have edge weights $w_{ij}$ along the edge $e_{ij} \coloneqq (v_i, v_j)$.

@@

We define the _extent_ of a simulation by the number of facilities that some patient has moved to or through within a given number of days. We see in @fig-extent-30 that we get starkly different distributions of extent for the naïve static model, compared to the other models.

@@

We can frame the empirical cumulative distribution (ECDF) of hitting times from a given seed hospital to a target hospital in a survival analysis view, and compare different experimental setups by comparing the ECDFs of the combinations of hitting times with log-rank tests. Each pair of hospitals yields a separate log-rank test statistic.
Using a crude Bonferroni correction to counteract the multiple comparisons (which occur due to the ${N_x}^2=338^2$ combinations of seed and target hospitals), we find that there are significant differences between each type of model. We note spikes at very low p-values when comparing the naïve static model to other models.

@@

fig, axs = plt.subplots(
    nrows=NN, ncols=NN, 
    squeeze=True, 
    sharex=True, sharey=True, 
    layout="constrained"
)

@@

    axs[index].axvline(confidence_threshold, color='r')
    axs[index].set_yscale('asinh')
    sns.despine(ax=axs[index])

@@

## Snapshots

Network analysis of the explicitly-temporal network is difficult, since the number of nodes is relatively large ($N \sim N_x \times N_t$), and inversely scales with the temporal discretisation chosen ($N_t \sim 1/W$).

We can instead look at the series of static network induced by projection onto the locations at each time point. This allows
